---
title: "MultiTraits_tutorial"
author: "He Yan"
date: "2025.10.1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MultiTraits_tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5)
```

## Introduction

MultiTraits is a comprehensive R package designed for advanced analysis and visualisation of multidimensional plant traits. It provides a standardised workflow and intuitive visualisation tools, streamlining the exploration of complex trait relationships for researchers, educators, and students. The package integrates four key analytical modules: Competitors-Stress tolerators-Ruderals (CSR) strategy analysis, Leaf-Height-Seed (LHS) strategy analysis, Niche Periodic Table (NPT) analysis, and Plant Trait Network (PTN) analysis. These modules offer complementary perspectives, promoting a more comprehensive understanding of multidimensional trait variation patterns in plants. This tutorial aims to guide users through the core functionalities and practical applications of the MultiTraits package. Through practical demonstrations, we illustrate how these analytical modules can be utilized to explore multidimensional relationships among plant traits, including key steps in data preparation, analysis workflows, and result visualization.

## Installation

### From CRAN (recommended)

```{r,class.source = 'fold-show'}
# install.packages("MultiTraits")
```

### From GitHub (development version)

```{r,class.source = 'fold-show'}
# if (!requireNamespace("devtools", quietly = TRUE)) {install.packages("devtools")}
# devtools::install_github("biodiversity-monitoring/MultiTraits")
```

## Main Functions

| Module         | Function                | Description                                                                                                                           |
|----------------|----------------|-----------------------------------------|
| **CSR Module** | `CSR()`                 | Classifies plant species' CSR ecological strategies based on leaf traits (LA, LDMC, and SLA) using the Pierce et al. (2017) method.   |
|                | `CSR_hodgson()`         | Implements the Hodgson et al. (1999) method for CSR classification.                                                                   |
|                | `CSR_plot()`            | Provides advanced ternary plot visualizations for CSR analysis, to explore plant adaptation to different environments.                |
| **LHS Module** | `LHS()`                 | Analyzes LHS strategies using SLA, plant height, and seed mass.                                                                       |
|                | `LHS_plot()`            | Visualizes three-dimensional trait distributions in LHS space.                                                                        |
|                | `LHS_strategy_scheme()` | Classifies strategies based on the LHS scheme.                                                                                        |
| **NPT Module** | `NPT_discrete()`        | Supports discrete niche classification schemes. Performs PCA and clustering (CART or K-means) to group species within each dimension. |
|                | `NPT_discrete_plot()`   | Visualizes the discrete niche classification scheme.                                                                                  |
|                | `NPT_continuous()`      | Supports continuous niche classification schemes. Performs a hierarchical PCA approach.                                               |
|                | `NPT_continuous_plot()` | Visualizes the results of the continuous niche classification.                                                                        |
| **PTN Module** | `PTN_corr()`            | Offers correlation computation and visualization for trait networks.                                                                  |
|                | `PTN()`                 | Generates trait correlation networks.                                                                                                 |
|                | `PTN_metrics()`         | Quantifies network metrics for trait networks.                                                                                        |
|                | `PTN_plot()`            | Visualizes the trait network graph.                                                                                                   |

### Example Datasets: PFF (Pine Forests Flora)

The package includes a built-in example dataset called PFF that can be loaded as follows:

```{r,class.source = 'fold-show'}
library(MultiTraits)
data(PFF)
# View the structure of the datasets
head(PFF)
```

The PFF dataset contains trait measurements for plant species found in pine forests, with the following variables:

Height: Maximum plant height (cm)

Leaf_area: Leaf area (mm²)

LDMC: Leaf dry matter content (%)

SLA: Specific leaf area (mm²/mg)

SRL: Specific root length

SeedMass: Seed mass

FltDate: Flowering date

FltDur: Flowering duration

k_value: Decomposition rate

Leaf_Cmass: Leaf carbon content

Leaf_Nmass: Leaf nitrogen content

Leaf_CN: Leaf carbon/nitrogen ratio

Leaf_Pmass: Leaf phosphorus content

Leaf_NP: Leaf nitrogen/phosphorus ratio

Leaf_CP: Leaf carbon/phosphorus ratio

Root_Cmass: Root carbon content

Root_Nmass: Root nitrogen content

Root_CN: Root carbon/nitrogen ratio

### CSR Strategy

The CSR (Competitor-Stress-Tolerator-Ruderal) strategy, initially conceptualised by Grime (1974), classifies plants based on their ecological strategies, which are influenced by environmental factors like competition, stress, and disturbance. The MultiTraits package offers two approaches for CSR analysis: the Hodgson et al. (1999) method and the Pierce et al. (2017) method. The Pierce method, used here, utilises leaf area (LA), leaf dry matter content (LDMC), and specific leaf area (SLA) to classify species. The CSR_plot() function visualises the results in a ternary plot, showing species' positions based on their relative C, S, and R strategy proportions. This aids in understanding plant functional classification and predicting responses to environmental changes.

```{r,class.source = 'fold-show'}
# Load packages
library(MultiTraits)
# Load the PFF dataset
data(PFF)
head(PFF)

# Create a data frame with the CSR traits
csr_traits <- data.frame(LA=PFF$Leaf_area, LDMC=PFF$LDMC, SLA=PFF$SLA)
rownames(csr_traits) <- PFF$species
head(csr_traits)

# Perform CSR analysis
csr_result <- CSR(csr_traits)
head(csr_result)

# Create the CSR plot
CSR_plot(csr_result, expand_margin = 1.05)
```


### LHS Strategy

The Leaf-Height-Seed (LHS) scheme, proposed by Westoby (1998), classifies plant species based on specific leaf area (SLA), canopy height at maturity, and seed mass. These three traits reflect trade-offs in resource acquisition, spatial occupation, and reproductive investment. The MultiTraits package implements the LHS scheme by positioning species in a three-dimensional trait space using log-transformed trait values. This continuous space is then divided into eight discrete strategy types (e.g., S-S-S, L-L-L) based on median trait values, reflecting different trade-offs in growth, competition, and survival. The LHS_plot() function visualises these strategies in 3D space, while LHS_strategy_scheme() provides a key for interpreting the strategy types.

```{r,class.source = 'fold-show'}
# Load packages
library(MultiTraits)
# Load the PFF dataset
data(PFF)
head(PFF)

# Create a data frame with the LHS traits
lhs_traits <- PFF[, c("SLA", "Height", "SeedMass")]
rownames(lhs_traits) <- PFF$species
head(lhs_traits)

# Perform LHS analysis
lhs_result <- LHS(lhs_traits)
head(lhs_result)
table(lhs_result$LHS_strategy)


# Create a visualization plot of the LHS analysis results
LHS_plot(lhs_result, group = "LHS_strategy", show_cube = TRUE)

# Display the LHS strategy scheme diagram
LHS_strategy_scheme()
```


### Niche Periodicity Table

The Niche Periodic Table (NPT) framework systematically describes and predicts species functions by identifying recurring trait combinations, driven by convergent evolution. Within the MultiTraits R package, the NPT module offers both discrete and continuous classification schemes. The NPT_discrete() function categorises species into distinct niche groups using PCA and clustering, with NPT_discrete_plot() visualising these results. For continuous classification, NPT_continuous() performs hierarchical PCA to create a two-dimensional ordination space, while NPT_continuous_plot() visualises these integrated niche patterns.

```{r,class.source = 'fold-show'}
# Load packages
library(MultiTraits)
# Load and prepare data
data(PFF)
rownames(PFF) <- PFF$species
PFF_traits <- PFF[, c("SLA", "SRL", "Leaf_Nmass", "Root_Nmass","Height",
                      "Leaf_CN", "Root_CN","SeedMass", "FltDate", "FltDur")]
# Perform log transformation of data and remove missing values
PFF_traits <- log(na.omit(PFF_traits))
head(PFF_traits)

# Define trait dimensions
dimension <- list(
  Grow = c("SLA", "SRL", "Leaf_Nmass", "Root_Nmass"),
  Survive = c("Height", "Leaf_CN", "Root_CN"),
  Reproductive = c("SeedMass", "FltDate", "FltDur")
)

# Perform discrete niche scheme
set.seed(123)
discrete_result <- NPT_discrete(data = PFF_traits, dimension = dimension)
discrete_result
NPT_discrete_plot(discrete_result$niche_classification)

# Perform continuous niche scheme
continuous_result <- NPT_continuous(data = PFF_traits, dimension = dimension)
continuous_result 
NPT_continuous_plot(continuous_result$result)
```


### Plant Trait Network

Plant Trait Networks (PTNs) provide a framework for understanding complex interdependencies among plant traits by treating traits as nodes and their associations as edges in a network. The PTN module, implemented as described by He et al. (2020), offers functions to calculate correlations (PTN_corr()), generate networks (PTN()), quantify network metrics (PTN_metrics()), and visualise networks (PTN_plot()). The package supports phylogenetic correction to account for species' shared evolutionary history. Significant correlations are selected based on thresholds, and network metrics, including node- and network-level parameters, are used to analyse trait interdependence.

```{r,class.source = 'fold-show'}
# Load packages
library(MultiTraits)
# Load and prepare data
data(PFF)
rownames(PFF) <- PFF$species
PFF_traits <- PFF[, c("Leaf_area","LDMC","SLA", "Leaf_Cmass","Leaf_Nmass",
                      "Leaf_CN","Leaf_Pmass", "Leaf_NP","Leaf_CP")]
# Perform log transformation of data and remove missing values
PFF_traits <- log(na.omit(PFF_traits))
head(PFF_traits)

data(PFF_tree)
PFF_tree

# Calculate trait correlations using specified thresholds
PTN_corr(traits_matrix=PFF_traits, rThres = 0.2, pThres = 0.05,method = "pearson",
         phylo_correction = TRUE,phylo_tree = PFF_tree)

# Perform Trait Network (TN) analysis
PTN_phylo_result <- PTN(traits_matrix = PFF_traits,
                      rThres = 0.2,
                      pThres = 0.05,
                      method = "pearson",
                      phylo_correction = TRUE,
                      phylo_tree = PFF_tree)

# Calculate network metrics for the trait network
PTN_metrics(PTN_phylo_result)

set.seed(22)
# Create visualization plots of the trait network
PTN_plot(PTN_phylo_result, style = 1, vertex.size = 20, vertex.label.cex = 0.6)
```

